---
title: "chla-indicators-2022"
author: "Matt Callahan"
date: '2023-08-08'
output: html_document
---

## Notes
In 2023 we switched chlorophyll products from MODIS to globcolour. 
I downloaded globcolour data and pushed it to the AKFIN database. 
That work can be found in the globcolour_download.R and globcolour_processing scripts.
Since the required data for ESP indicators is in AKFIN I will directly push to the database. 
This document meerly shows checks to make sure I get the same results as before when I just did it in R.


## AMJ_Chlorophylla_Biomass_SEBS_Satellite
Confirm that database pull matches R calculations. Yep! Looks low though...

```{r}
library(tidyverse)
library(lubridate)
library(odbc)
library(getPass)
#connect to akfin
con <- dbConnect(odbc::odbc(), "akfin", UID=getPass(msg="USER NAME"), PWD=getPass()) 

#recalculate indicator for 2023 from the akfin database
glob_AMJ_SEBS <-dbFetch(dbSendQuery(con, "select extract(year from to_date(start_date,'YYYY-MM-DD')) year, 'AMJ_Chlorophylla_Biomass_SEBS_Satellite' as indicator_name, round(avg(chla),2) data_value
from env_data.globcolour_2023 a
left join env_data.globcolour_spatial_lookup b on a.glob_id=b.glob_id
where extract(month from to_date(start_date,'YYYY-MM-DD')) in (4, 5, 6)
and bsierp_region_name in ('St. Matthew','Pribilofs')
group by extract(year from to_date(start_date,'YYYY-MM-DD'))
order by extract(year from to_date(start_date,'YYYY-MM-DD')) asc"))
  
#calulcate from RDS files
glob_AMJ_SEBS2 <- readRDS("../../data/2023_combined/2023_for_akfin.RDS") %>%
  left_join(readRDS("../../data/globcolour_spatial_lookup.RDS"),
            by="glob_id")%>%
  mutate(month=month(ymd(start_date))) %>%
  filter(bsierp_region_name%in%c("St. Matthew",
                          "Pribilofs") & 
           month%in%c(4:6)) %>%
  summarize(data_value=mean(chla), n_values=n())

 
```


## Spring_Chlorophylla_Biomass_SEBS_Inner_Shelf_Satellite
```{r}
#recalculate indicator for 2023 from the akfin database
glob_BBRKC <-dbFetch(dbSendQuery(con, "select extract(year from to_date(start_date,'YYYY-MM-DD')) year, 'Spring_Chlorophylla_Biomass_SEBS_Inner_Shelf_Satellite' as indicator_name, round(avg(chla),2) data_value
from env_data.globcolour_2023 a
left join env_data.globcolour_spatial_lookup b on a.glob_id=b.glob_id
where extract(month from to_date(start_date,'YYYY-MM-DD')) in (4, 5, 6)
and bs_king='BBRKC'
group by extract(year from to_date(start_date,'YYYY-MM-DD'))
order by extract(year from to_date(start_date,'YYYY-MM-DD')) asc"))
  
#calulcate from RDS files
glob_BBRCK2 <- readRDS("../../data/2023_combined/2023_for_akfin.RDS") %>%
  left_join(readRDS("../../data/globcolour_spatial_lookup.RDS"),
            by="glob_id")%>%
  mutate(month=month(ymd(start_date))) %>%
  filter(bs_king =='BBRKC' & 
           month%in%c(4:6)) %>%
  summarize(data_value=mean(chla), n_values=n())

```

## Spring_Chlorophylla_Biomass_SMBKC_Satellite

```{r}
glob_stmatt <-dbFetch(dbSendQuery(con, "select extract(year from to_date(start_date,'YYYY-MM-DD')) year, 'Spring_Chlorophylla_Biomass_SMBKC_Satellite' as indicator_name, round(avg(chla),2) data_value
from env_data.globcolour_2023 a
left join env_data.globcolour_spatial_lookup b on a.glob_id=b.glob_id
where extract(month from to_date(start_date,'YYYY-MM-DD')) in (4, 5, 6)
and bs_king='StMattBKC'
group by extract(year from to_date(start_date,'YYYY-MM-DD'))
order by extract(year from to_date(start_date,'YYYY-MM-DD')) asc"))
  
#calulcate from RDS files
glob_stmatt2 <- readRDS("../../data/2023_combined/2023_for_akfin.RDS") %>%
  left_join(readRDS("../../data/globcolour_spatial_lookup.RDS"),
            by="glob_id")%>%
  mutate(month=month(ymd(start_date))) %>%
  filter(bs_king =='StMattBKC' & 
           month%in%c(4:6)) %>%
  summarize(data_value=mean(chla), n_values=n())


```

## Groundfish indicators
Here I calculate 2022 groundfish indicators for the GOA. Jens Nielson will calculate the BS

Spring_Chlorophylla_Peak_EGOA_Satellite
Spring_Chlorophylla_Peak_WCGOA_Satellite
Spring_Chlorophylla_Biomass_EGOA_Satellite
Spring_Chlorophylla_Biomass_WCGOA_Satellite


## Spring_Chlorophylla_Biomass_GOA_Satellite


```{r}
#define indicator

glob_goa <-dbFetch(dbSendQuery(con, "select extract(year from to_date(start_date,'YYYY-MM-DD')) year, 'Spring_Chlorophylla_Biomass_GOA_Satellite' as indicator_name, round(avg(chla),2) data_value
from env_data.globcolour_2023 a
left join env_data.globcolour_spatial_lookup b on a.glob_id=b.glob_id
where extract(month from to_date(start_date,'YYYY-MM-DD')) in (4, 5, 6)
and NMFS_REP_AREA in ('610', '620', '630', '640', '650')
and depth <= (-10)
group by extract(year from to_date(start_date,'YYYY-MM-DD'))
order by extract(year from to_date(start_date,'YYYY-MM-DD')) asc"))
  
#calulcate from RDS files
glob_goa2 <- readRDS("../../data/2023_combined/2023_for_akfin.RDS") %>%
  left_join(readRDS("../../data/globcolour_spatial_lookup.RDS"),
            by="glob_id")%>%
  mutate(month=month(ymd(start_date))) %>%
  filter(nmfs_rep_area %in% c('610', '620', '630', '640', '650') & 
           month%in%c(4,5,6) &
           depth<= (-10)) %>%
  summarize(data_value=mean(chla), n_values=n())

```

## Spring_Chlorophylla_Peak_GOA_Satellite

This was originally Spring_Chlorophylla_Biomass_EGOA_Satellite. At the request of the sablefish ESP team we expanded the extent to gulf-wide and removed the depth filter.

I am just going to submit these the old fashion way...

```{r}
#define indicator
glob_goa_pk <-dbFetch(dbSendQuery(con, "WITH ranked_data AS (
    SELECT
         year,
         doy,
        meanchla,
        ROW_NUMBER() OVER (PARTITION BY year ORDER BY meanchla DESC) AS rn
        from (select extract(year from to_date(start_date,'YYYY-MM-DD')) year,
        to_number(to_char(to_date(start_date,'YYYY-MM-DD'),'DDD')) doy, 
        round(avg(chla),2) meanchla 
    FROM env_data.globcolour_2023 a
left join env_data.globcolour_spatial_lookup b on a.glob_id=b.glob_id
where extract(month from to_date(start_date,'YYYY-MM-DD')) in (4, 5, 6)
and NMFS_REP_AREA in ('610', '620', '630', '640', '650')
and depth <= (-10)
        group by  extract(year from to_date(start_date,'YYYY-MM-DD')),
        to_number(to_char(to_date(start_date,'YYYY-MM-DD'),'DDD'))
))
SELECT
    year,
    'Spring_Chlorophylla_Peak_GOA_Satellite' as indicator_name,
    doy data_value
FROM ranked_data
WHERE rn = 1"))

#calculate indicator value
glob_goa_pk2 <- readRDS("../../data/2023_combined/2023_for_akfin.RDS") %>%
  left_join(readRDS("../../data/globcolour_spatial_lookup.RDS"),
            by="glob_id") %>%
  mutate(month=month(ymd(start_date))) %>%
  filter(nmfs_rep_area %in% c(610, 620, 630, 640, 650) & month %in% c(4:6) & depth<=(-10)) %>%
  group_by(start_date) %>%
  summarise(WEEKLY_VALUE=mean(chla,na.rm=TRUE))%>%
  slice(which.max(WEEKLY_VALUE))%>%
  mutate(DATA_VALUE=yday(start_date))


```




## Spring_Chlorophylla_Biomass_WCGOA_Satellite

I used the same filters Jordan included in his github code, the values are close, but do not match perfectly.

```{r, warning=FALSE, message=FALSE}
glob_wcgoa <-dbFetch(dbSendQuery(con, "select extract(year from to_date(start_date,'YYYY-MM-DD')) year, 'Spring_Chlorophylla_Biomass_WCGOA_Satellite' as indicator_name, round(avg(chla),2) data_value, count(*) n_values
from env_data.globcolour_2023 a
left join env_data.globcolour_spatial_lookup b on a.glob_id=b.glob_id
where extract(month from to_date(start_date,'YYYY-MM-DD')) = 5 
and NMFS_REP_AREA in ('610', '620', '630')
and depth <= (-10)
and depth > (-200)
group by extract(year from to_date(start_date,'YYYY-MM-DD'))
order by extract(year from to_date(start_date,'YYYY-MM-DD')) asc
"))
  
#calulcate from RDS files
glob_wcgoa2 <- readRDS("../../data/2023_combined/2023_for_akfin.RDS") %>%
  left_join(readRDS("../../data/globcolour_spatial_lookup.RDS"),
            by="glob_id")%>%
  mutate(month=month(ymd(start_date))) %>%
  filter(nmfs_rep_area %in% c('610', '620', '630') & 
           month%in%c(5) &
           depth <= (-10) &
           depth > (-200)) %>%
  summarize(data_value=mean(chla), n_values=n())


```

## Spring_Chlorophylla_Peak_WCGOA_Satellite

```{r}

glob_wcgoa_pk <-dbFetch(dbSendQuery(con, "WITH ranked_data AS (
    SELECT
         year,
         doy,
        meanchla,
        ROW_NUMBER() OVER (PARTITION BY year ORDER BY meanchla DESC) AS rn
        from (select extract(year from to_date(start_date,'YYYY-MM-DD')) year,
        to_number(to_char(to_date(start_date,'YYYY-MM-DD'),'DDD')) doy, 
        round(avg(chla),2) meanchla 
    FROM env_data.globcolour_2023 a
left join env_data.globcolour_spatial_lookup b on a.glob_id=b.glob_id
where extract(month from to_date(start_date,'YYYY-MM-DD')) in (4, 5, 6)
and NMFS_REP_AREA in ('610', '620', '630')
and depth <= (-10)
and depth >= (-200)
        group by  extract(year from to_date(start_date,'YYYY-MM-DD')),
        to_number(to_char(to_date(start_date,'YYYY-MM-DD'),'DDD'))
))
SELECT
    year,
    'Spring_Chlorophylla_Peak_GOA_Satellite' as indicator_name,
    doy data_value
FROM ranked_data
WHERE rn = 1"))

glob_wcgoa_pk <- readRDS("../../data/2023_combined/2023_for_akfin.RDS") %>%
  left_join(readRDS("../../data/globcolour_spatial_lookup.RDS"),
            by="glob_id") %>%
  mutate(month=month(ymd(start_date))) %>%
  filter(nmfs_rep_area %in% c('610', '620', '630') & month %in% c(4:6) & depth<=(-10) & depth>=(-200)) %>%
  group_by(start_date) %>%
  summarise(WEEKLY_VALUE=mean(chla,na.rm=TRUE))%>%
  slice(which.max(WEEKLY_VALUE))%>%
  mutate(DATA_VALUE=yday(start_date))
```

## quickly plot chla indicators
```{r}


#combine data
glob_indicators<-glob_goa %>%
  bind_rows(glob_wcgoa) %>%
  bind_rows(glob_AMJ_SEBS) %>%
  bind_rows(glob_BBRKC) %>%
  bind_rows(glob_stmatt)

#pull viirs indicators from chl-indicator-comparison/2023_viirs_esp_indicators.R
viirs<-readRDS("data/viirs_indicators_2023.RDS")


plotfun<-function(indy) {
  ggplot()+
    geom_line(data=glob_indicators%>%filter(INDICATOR_NAME==indy),
              aes(x=YEAR, y=DATA_VALUE, color="black"))+
    geom_line(data=viirs%>%filter(indicator_name==indy),
              aes(x=year, y=mean_viirs, color="red"))+
    scale_color_manual(name= "",values=c("black"="black", "red"="red"), labels=c("globcolour","viirs"))+
    ggtitle(paste0(indy))+
    theme_bw()
    
}

  
indicators<-unique(glob_indicators$INDICATOR_NAME)

for (indy in indicators) {
 plot <- plotfun(indy)
  png_file <- paste0("ESP/2023_indicators/qaplots/", indy, ".png")
  ggsave(filename = png_file, plot = plot, width = 6, height = 4, units = "in")
}

```

