---
title: "chla-indicators-2022"
author: "Matt Callahan"
date: '2023-08-08'
output: html_document
---

## Notes
In 



## AMJ_Chlorophylla_Biomass_SEBS_Satellite
Confirm that database pull matches R calculations. Yep! Looks low though...

```{r}
library(tidyverse)
library(lubridate)
library(odbc)
library(getPass)
#connect to akfin
con <- dbConnect(odbc::odbc(), "akfin", UID=getPass(msg="USER NAME"), PWD=getPass()) 
#define indicator
indicator<-"AMJ_Chlorophylla_Biomass_SEBS_Satellite"
#MODIS

#recalculate indicator for 2023 from the akfin database
glob_AMJ_SEBS <-dbFetch(dbSendQuery(con, "select round(avg(chla),2) data_value, count(*) n_values 
from globcolour_2023 a
left join globcolour_spatial_lookup b on a.glob_id=b.glob_id
where extract(year from to_date(start_date,'YYYY-MM-DD'))=2023
and extract(month from to_date(start_date,'YYYY-MM-DD')) in (4, 5, 6)
and bsierp_region_name in ('St. Matthew','Pribilofs')"))
  
#calulcate from RDS files
glob_AMJ_SEBS2 <- readRDS("../../data/2023_combined/2023_for_akfin.RDS") %>%
  left_join(readRDS("../../data/globcolour_spatial_lookup.RDS"),
            by="glob_id")%>%
  mutate(month=month(ymd(start_date))) %>%
  filter(bsierp_region_name%in%c("St. Matthew",
                          "Pribilofs") & 
           month%in%c(4:6)) %>%
  summarize(data_value=mean(chla), n_values=n())

#recalculate indicator
mod_AMJ_SEBS2 <- mod_ebs2 %>%
  mutate(YEAR=year)%>%
  filter(BSIERP_Region_Name%in%c("St. Matthew",
                          "Pribilofs"
                          ) & 
           month%in%c(4:6) 
         ) %>%
  group_by(YEAR) %>%
  summarise(MODIS_VALUE=mean(chlorophyll,na.rm=TRUE))#%>%

 
```

The following chunk calculates the indicator for 2022, binds to the 2003-2021 data, and saves as a csv.
```{r}
#recalculate indicator
mod_AMJ_SEBS_2022 <- mod2022 %>%
  mutate(YEAR=year)%>%
  filter(BSIERP_Region_Name%in%c("St. Matthew",
                          "Pribilofs"
                          ) & 
           month%in%c(4:6) 
         ) %>%
  group_by(YEAR) %>%
  summarise(MODIS_VALUE=mean(chlorophyll,na.rm=TRUE))

mod_AMJ_SEBS2<-mod_AMJ_SEBS2%>%
  bind_rows(mod_AMJ_SEBS_2022)

mod_AMJ_SEBS2$INDICATOR_NAME<-indicator
mod_AMJ_SEBS2<-mod_AMJ_SEBS2%>%
  mutate(DATA_VALUE=MODIS_VALUE)%>%
  dplyr::select(c(YEAR, INDICATOR_NAME, DATA_VALUE))

write.csv(mod_AMJ_SEBS2, "2022_indicators/AMJ_CHLA_SEBS.csv", row.names=FALSE)
```



## Spring_Chlorophylla_Biomass_SEBS_Inner_Shelf_Satellite
I'm just going to calculate these 2.
```{r}
#define indicator
indicator<-"Spring_Chlorophylla_Biomass_SEBS_Inner_Shelf_Satellite"
#MODIS
#recalculate indicator
mod_BB <- mod_ebs2 %>%
  mutate(YEAR=year) %>%
  filter(BS_KING=="BBRKC" & month%in%c(4:6)) %>%
  group_by(YEAR) %>%
  summarise(DATA_VALUE=mean(chlorophyll,na.rm=TRUE))%>%
  mutate(INDICATOR_NAME=indicator)%>%
  dplyr::select(c(YEAR, INDICATOR_NAME, DATA_VALUE))

#recalculate for 2022
bb22 <- mod2022 %>%
  mutate(YEAR=year)%>%
  filter(BS_KING=="BBRKC" & month%in%c(4:6)) %>%
  group_by(YEAR) %>%
  summarise(DATA_VALUE=mean(chlorophyll,na.rm=TRUE))%>%
  mutate(INDICATOR_NAME=indicator)%>%
  dplyr::select(c(YEAR, INDICATOR_NAME, DATA_VALUE))

mod_BB<-mod_BB%>%
  bind_rows(bb22)

write.csv(mod_BB, "2022_indicators/SPRING_CHLA_SEBS_IS.csv", row.names=FALSE)
```

## Spring_Chlorophylla_Biomass_SMBKC_Satellite

```{r}
#define indicator
indicator<-"Spring_Chlorophylla_Biomass_SMBKC_Satellite"
#MODIS
#recalculate indicator
mod_smbkc <- mod_ebs2 %>%
  mutate(YEAR=year) %>%
  filter(BS_KING=="StMattBKC" & month%in%c(4:6)) %>%
  group_by(YEAR) %>%
  summarise(DATA_VALUE=mean(chlorophyll,na.rm=TRUE))%>%
  mutate(INDICATOR_NAME=indicator)%>%
  dplyr::select(c(YEAR, INDICATOR_NAME, DATA_VALUE))

#recalculate indicator for 2022
sm22 <- mod2022 %>%
  mutate(YEAR=year) %>%
  filter(BS_KING=="StMattBKC" & month%in%c(4:6)) %>%
  group_by(YEAR) %>%
  summarise(DATA_VALUE=mean(chlorophyll,na.rm=TRUE))%>%
  mutate(INDICATOR_NAME=indicator)%>%
  dplyr::select(c(YEAR, INDICATOR_NAME, DATA_VALUE))

mod_smbkc<-mod_smbkc%>%
  bind_rows(sm22)

write.csv(mod_smbkc, "2022_indicators/SPRING_CHLA_SMBKC.csv", row.names=FALSE)
```

## Groundfish indicators
Here I calculate 2022 groundfish indicators for the GOA. Jens Nielson will calculate the BS

Spring_Chlorophylla_Peak_EGOA_Satellite
Spring_Chlorophylla_Peak_WCGOA_Satellite
Spring_Chlorophylla_Biomass_EGOA_Satellite
Spring_Chlorophylla_Biomass_WCGOA_Satellite

 
##GOA    

```{r}
#pull bering sea
#function to filter EBS
chla_goa<-function(x){
  x%>%filter(NMFS_REP_AREA %in% c(610, 620, 630, 640, 650))
}

#bind rows with other years
mod_goa<-lapply(2003:2022, function(x) readRDS(paste0("Data/MODIS/ESR/mod_",x,"_strata.RDS"))%>%
                     chla_goa())%>%
  bind_rows()
```



## Spring_Chlorophylla_Biomass_GOA_Satellite

This was originally Spring_Chlorophylla_Biomass_EGOA_Satellite. At the request of the sablefish ESP team we expanded the extent to gulf-wide and removed the depth filter.

```{r}
#define indicator
indicator<-"Spring_Chlorophylla_Biomass_GOA_Satellite"

#calculate indicator value
mod_SPR_GOA_B <- mod_goa %>%
  mutate(YEAR=year) %>%
  filter(NMFS_REP_AREA %in% c(610, 620, 630, 640, 650) & month %in% c(4:6) & depth<(-10)) %>%
  group_by(YEAR) %>%
  summarise(DATA_VALUE=mean(chlorophyll,na.rm=TRUE))%>%
  mutate(INDICATOR_NAME=indicator)%>%
  dplyr::select(c(YEAR, INDICATOR_NAME, DATA_VALUE))

write.csv(mod_SPR_GOA_B, "2022_indicators/Spring_Chlorophylla_Biomass_GOA_Satellite.csv", row.names=FALSE)
```

## Spring_Chlorophylla_Peak_GOA_Satellite

This was originally Spring_Chlorophylla_Biomass_EGOA_Satellite. At the request of the sablefish ESP team we expanded the extent to gulf-wide and removed the depth filter.

```{r}
#define indicator
indicator<-"Spring_Chlorophylla_Peak_GOA_Satellite"

#calculate indicator value
mod_SPR_GOA_P <- mod_goa %>%
  filter(NMFS_REP_AREA %in% c(610, 620, 630, 640, 650) & month %in% c(4:6) & depth<=(-10)) %>%
  group_by(date) %>%
  summarise(WEEKLY_VALUE=mean(chlorophyll,na.rm=TRUE))%>%
  mutate(YEAR=year(date))%>%
  group_by(YEAR)%>%
  slice(which.max(WEEKLY_VALUE))%>%
  mutate(DATA_VALUE=yday(date))%>%
  mutate(INDICATOR_NAME=indicator)%>%
  dplyr::select(c(YEAR, INDICATOR_NAME, DATA_VALUE))

write.csv(mod_SPR_GOA_P, "2022_indicators/Spring_Chlorophylla_Peak_GOA_Satellite.csv", row.names=FALSE)
```


## Spring_Chlorophylla_Biomass_WCGOA_Satellite

I used the same filters Jordan included in his github code, the values are close, but do not match perfectly.

```{r, warning=FALSE, message=FALSE}
#define indicator
indicator<-"Spring_Chlorophylla_Biomass_WCGOA_Satellite"

#calculate indicator value
mod_SPR_WCGOA_B <- mod_goa %>%
  mutate(YEAR=year) %>%
  filter(NMFS_REP_AREA %in% c(610, 620, 630) & month==5 & depth<(-10) & depth>(-200)) %>%
  group_by(YEAR) %>%
  summarise(DATA_VALUE=mean(chlorophyll,na.rm=TRUE))%>%
  mutate(INDICATOR_NAME=indicator)%>%
  dplyr::select(c(YEAR, INDICATOR_NAME, DATA_VALUE))

#write csv
write.csv(mod_SPR_WCGOA_B, "2022_indicators/Spring_Chlorophylla_Biomass_WCGOA_Satellite.csv", row.names=FALSE)

```

## Spring_Chlorophylla_Peak_WCGOA_Satellite

```{r}
#define indicator
indicator<-"Spring_Chlorophylla_Peak_WCGOA_Satellite"
#MODIS
#recalculate indicator
mod_SPR_WCGOA_P <- mod_goa %>%
  filter(NMFS_REP_AREA %in% c(610, 620, 630) & month %in% c(4:6) & depth<=(-10) & depth>=(-200)) %>%
  group_by(date) %>%
  summarise(WEEKLY_VALUE=mean(chlorophyll,na.rm=TRUE))%>%
  mutate(YEAR=year(date))%>%
  group_by(YEAR)%>%
  slice(which.max(WEEKLY_VALUE))%>%
  mutate(DATA_VALUE=yday(date))%>%
  mutate(INDICATOR_NAME=indicator)%>%
  dplyr::select(c(YEAR, INDICATOR_NAME, DATA_VALUE))


#write csv
write.csv(mod_SPR_WCGOA_P, "2022_indicators/Spring_Chlorophylla_Peak_WCGOA_Satellite.csv", row.names=FALSE)
```

## quickly plot chla indicators
```{r}
ggplot()+
  geom_line(data=mod_SPR_WCGOA_P, aes(x=YEAR, y=DATA_VALUE))+
  geom_line(data=mod_SPR_GOA_P, aes(x=YEAR, y=DATA_VALUE), color="red")+
  ggtitle("Peak chla: red=GOA, black=WCGOA")

ggplot()+
  geom_line(data=mod_SPR_WCGOA_B, aes(x=YEAR, y=DATA_VALUE))+
  geom_line(data=mod_SPR_GOA_B, aes(x=YEAR, y=DATA_VALUE), color="red")+
  ggtitle("Chla biomass: red=GOA, black=WCGOA")
```

