---
title: "ESR-Data_2022"
author: "Matt Callahan"
date: '2022-09-06'
output: html_document
---

## Overview

For 2022 the ESRs and ESPs, we download 8 day MODIS and append to previously created lookup tables. This code was originally saved as https://github.com/MattCallahan-NOAA/chla-indicator-comparison/blob/main/ESR-data-2022.Rmd and included OC-CCI downloads as well. Note: Do to large file sizes, data are not posted on github. 

1) download MODIS data (by year, 
east and west separate)
2) Convert to tidy objects
3) Join East and West for each year
4) Create MODIS lookup table
5) Join lookup table fields for each year
6) Distribute data to Noel/Jens/Jeanette

## MODIS

**download data**

```{r}
#load packages
library(tidync)
library(tidyverse)
library(lubridate)
require(sf)
require(AKmarineareas)
library(marmap)

#set long timeout
options(timeout=6000)

#Download data
myyear <- 2003:2021 # I already downloaded 2020-2022
#East
for(i in myyear){
  file_name <- paste0("Data/MODIS/ESR/mod_",i,"_e.nc")
  download.file(url = paste0("https://coastwatch.pfeg.noaa.gov/erddap/griddap/erdMH1chla8day.nc?chlorophyll%5B(",
                             i,"-01-01T12:00:00Z):(", i,"-12-31T12:00:00Z)%5D%5B(69.02083):(46.02083)%5D%5B(-179.9792):(-129.9792)%5D&.draw=surface&.vars=longitude%7Clatitude%7CsstMasked&.colorBar=%7C%7C%7C%7C%7C&.bgColor=0xffccccff"),
                method = "libcurl", mode="wb",destfile = file_name)
}

#west
for(i in myyear){
  file_name <- paste0("Data/MODIS/ESR/mod_",i,"_w.nc")
  download.file(url = paste0("https://coastwatch.pfeg.noaa.gov/erddap/griddap/erdMH1chla8day.nc?chlorophyll%5B(",
                             i,"-01-01T12:00:00Z):(", i,"-12-31T12:00:00Z)%5D%5B(56.5):(46.5)%5D%5B(167):(179.9792)%5D&.draw=surface&.vars=longitude%7Clatitude%7CsstMasked&.colorBar=%7C%7C%7C%7C%7C&.bgColor=0xffccccff"),
                method = "libcurl", mode="wb",destfile = file_name)
}

#2022
#east of dateline
download.file(url = "https://coastwatch.pfeg.noaa.gov/erddap/griddap/erdMH1chla8day.nc?chlorophyll%5B(2022-01-01T12:00:00Z):(2022-06-14T12:00:00Z)%5D%5B(69.02083):(46.02083)%5D%5B(-179.9792):(-129.9792)%5D&.draw=surface&.vars=longitude%7Clatitude%7CsstMasked&.colorBar=%7C%7C%7C%7C%7C&.bgColor=0xffccccff", method = "libcurl", mode="wb",destfile = "Data/MODIS/ESR/mod_2022_e.nc")
#west
download.file(url = "https://coastwatch.pfeg.noaa.gov/erddap/griddap/erdMH1chla8day.nc?chlorophyll%5B(2022-01-01T12:00:00Z):(2022-06-14T12:00:00Z)%5D%5B(56.5):(46.5)%5D%5B(167):(179.9792)%5D&.draw=surface&.vars=longitude%7Clatitude%7CsstMasked&.colorBar=%7C%7C%7C%7C%7C&.bgColor=0xffccccff", method = "libcurl", mode="wb",destfile = "Data/MODIS/ESR/mod_2022_w.nc")

```

Unfortunately April 11 was blank and April 19 was missing. One day MODIS was missing a lot of values from this period. We will add June to the original 2022 but April will be missing. from this year's ESR contribution. This code loads the last two weeks in June to be added to the 2022 files downloaded above.

```{r}
#2022 ceased to be updated but a new url was provided for us erdMH1chla8day_R2022NRT use this for June 22-30 2022

#east of dateline
download.file(url = "https://coastwatch.pfeg.noaa.gov/erddap/griddap/erdMH1chla8day_R2022NRT.nc?chlorophyll%5B(2022-06-22T12:00:00Z):(2022-06-30T12:00:00Z)%5D%5B(69.02083):(46.02083)%5D%5B(-179.9792):(-129.9792)%5D&.draw=surface&.vars=longitude%7Clatitude%7CsstMasked&.colorBar=%7C%7C%7C%7C%7C&.bgColor=0xffccccff", method = "libcurl", mode="wb",destfile = "Data/MODIS/ESR/mod_jun_2022_e.nc")
#west
download.file(url = "https://coastwatch.pfeg.noaa.gov/erddap/griddap/erdMH1chla8day_R2022NRT.nc?chlorophyll%5B(2022-06-22T12:00:00Z):(2022-06-30T12:00:00Z)%5D%5B(56.5):(46.5)%5D%5B(167):(179.9792)%5D&.draw=surface&.vars=longitude%7Clatitude%7CsstMasked&.colorBar=%7C%7C%7C%7C%7C&.bgColor=0xffccccff", method = "libcurl", mode="wb",destfile = "Data/MODIS/ESR/mod_jun_2022_w.nc")


```

**convert to tidy data objects**

```{r}
#function to convert to data frame
tidy_chl<-function(file) {
  tidync(file) %>% 
  hyper_tibble()%>% 
  mutate(date=as_datetime(time),
         lonc=as.numeric(ifelse(longitude<0, #ifelse statement keeps +- lons the same length
                                substr(longitude,1,8),
                                substr(longitude,1,7))), 
         latc=as.numeric(substr(latitude,1,6)),
         chlorophyll=round(chlorophyll,3),
         year=year(date),
         month=month(date))
}

#Merge east and west for each year and save as RDS
lapply(2003:2022, function(x) tidy_chl(paste0("Data/MODIS/ESR/mod_",x,"_e.nc"))%>%
               bind_rows(tidy_chl(paste0("Data/MODIS/ESR/mod_",x,"_w.nc")))%>%
         saveRDS(paste0("Data/MODIS/ESR/mod_",x,".RDS")))

rm(mod_e, mod_w)

mod_2003<-readRDS("Data/MODIS/ESR/mod_2003.RDS")

ggplot()+
  geom_point(data=mod_2003%>%filter(month==6), aes(x=ifelse(lonc>0, lonc, lonc+360), y=latc))+
  geom_sf(data=ai%>%st_shift_longitude(), color="red", fill=NA)
   
```


## Assign spatial strata
The existing lookup table needs to be updated to reflect recent changes. I will make a new lookup table as I did for VIIRS and OCCCI

**create grid**
The first step is to create the lookup grid that can be joined with chla data.

```{r}
#create grid
#determine extents
mod_2003%>% filter(lonc<0) %>% summarise(MIN=min(lonc), MAX=max(lonc))
mod_2003%>% filter(lonc>0) %>% summarise(MIN=min(lonc), MAX=max(lonc))
min(mod_2003$latc); max(mod_2003$latc)

lon<-as.numeric(substr(head(sort(unique(mod_2003$longitude)), 21), 1, 9))
lon
#create new vectors with 4 decimal places
lon_lkp<-c(as.numeric(substr(seq(from=-179.97916666666666667, to=-129.979, by=0.0416666666666667), 1, 8)), #east
          as.numeric(substr(seq(from=166.97916666666666667, to=179.9792, by=0.0416666666666667), 1, 7))) #west
lat_lkp<-as.numeric(substr(seq(from=46.97916666666666667, to=69.0209, by=0.0416666666666667), 1, 7))

#create grid
modis_4k_grid<-expand.grid(lon_lkp, lat_lkp)%>%
           mutate(lonc=as.numeric(ifelse(Var1<0, 
                                substr(Var1,1,8),
                                substr(Var1,1,7))), #from Jordan's code
         latc=as.numeric(substr(Var2,1,6)))%>%
  rename(latitude=Var2, longitude=Var1)

#test join
test_join<-mod_2003%>%
  inner_join(modis_4k_grid, by=c("lonc"="lonc", "latc"="latc")) 

ggplot()+
      geom_point(data=test_join%>%filter(month==6), aes(x=ifelse(longitude.y>0, longitude.y, longitude.y+360), y=latitude.y), color="red")+
  geom_point(data=test_join%>%filter(month==6), aes(x=ifelse(longitude.x>0, longitude.x, longitude.x+360), y=latitude.x))

  
```

**Add marine area fields**
We need the same fields as in the crw sst spatial lookup table. 

*statefed* Flags if the location is in "state", "fed" or "intl" (international) waters.    

*stat_area* ADF&G statistical area (there are 1758 of these total) <http://www.adfg.alaska.gov/index.cfm?adfg=fishingCommercialByFishery.statmaps>    

*nmfsarea*  NMFS management areas <https://www.fisheries.noaa.gov/alaska/sustainable-fisheries/alaska-fisheries-figures-maps-boundaries-regulatory-areas-and-zones>   

    

*Ecosystem*  From the AK_MES shapefile created for the ESRs. The shapefiles has four areas but we omitted Arctic.    

*Ecosystem_sub*  The AK_MES shapefile did not include all of the ESR subregions so I created them. NBS / SEBS are split at 60. EGOA and WGOA are split at 144. The eastern and western Aleutians are split at -177 and positive 177, with the central region in between.  

*bsierp_name* and *bsierp_id*  Name and ID of BSIERP area.

*crab areas* Bering Sea crab management areas: *nbs_crab*, *bs_king*, and *bs_tanner*.

*depth*  This is the GEBCO bathymetry dataset as pulled by the marmap package using resolution=1. Data have been filtered for depths<0  

We will use ADFG for our spatial extent. 1) they cover the rest of the ecosystem, NMFS, and BSIERP regions except the arctic, 2) The CRW lookup table uses the same extent, and 3) Spatial aggregations by stat area may become an importan seconday product as this data is gap-rich. 

**Download sptial data**
Spatial Data were loaded from two esri file geodatabases prepared by Brett Holycross at PSMFC. The Alaska Marine Management Areas layers contain ADFG, NMFS, and BSIERP regions, that have long been static. Ecosystem and Ecosystem subareas reflect the updated boundaries requested by Ecosystem Status Report Authors in 2022. The Bering Sea Crab area layers were created in consultation with Erin Fedewa (NMFS).

```{r}
#add eval=FALSE before knitting
#load spatial data
#marine management areas
Management <- st_read(dsn="Data/GIS/Alaska_Marine_Management_Areas.gdb", layer="Alaska_Marine_Areas_dd")
unique(Management$Area_Type)
ADFG<-Management%>%filter(Area_Type=="ADFG Stat Area")%>%
  dplyr::select(STAT_AREA, WATERS_COD, Shape_Length, Shape_Area, Shape)
NMFS<-Management%>%filter(Area_Type=="NMFS Reporting Area")%>%
  dplyr::select(NMFS_REP_AREA, Shape_Length, Shape_Area, Shape)
ESR<-Management%>%filter(Area_Type=="Ecosystem Subarea")%>%
  dplyr::select(Ecosystem_Subarea, Shape_Length, Shape_Area, Shape)
LME<-Management%>%filter(Area_Type=="Ecosystem Area")%>%
  dplyr::select(Ecosystem_Area, Shape_Length, Shape_Area, Shape)
BSIERP<-Management%>%filter(Area_Type=="BSIERP Region")%>%
  dplyr::select(BSIERP_ID, BSIERP_Region_Name, Shape_Length, Shape_Area, Shape)
#crab areas
nbs_crab <- st_read(dsn="Data/GIS/AK_Crab_Management_Areas.gdb", layer="NBS_Crab_dd")
bs_tanner_crab <- st_read(dsn="Data/GIS/AK_Crab_Management_Areas.gdb", layer="BS_Tanner_Crab_dd")
bs_king_crab <- st_read(dsn="Data/GIS/AK_Crab_Management_Areas.gdb", layer="BS_King_Crab_dd")

```

Add sptatial data to lookup table

**Add spatial components to lookup table**
```{r, eval=FALSE}
#Turn off spherical geometry
sf::sf_use_s2(FALSE)

#point in polygon to add regional information to grid
#convert grid to spatial object
grid <- modis_4k_grid%>%
  st_as_sf(coords = c('longitude', 'latitude'), crs = 4326, agr = 'constant')

#ADFG
grid<-st_join(grid, ADFG, join = st_within)

#plot
ggplot()+
  geom_sf(data=grid, aes(color=STAT_AREA))

#limit to points with a stat area
#brings from >1 million to <400k
grid<-grid%>%
  filter(STAT_AREA>0)

#NMFS
grid<-grid%>%
  st_join(NMFS, join = st_within)

#way slow. Save output 
saveRDS(grid, "Data/MODIS/grid_adfg_nmfs.RDS")

#Ecosystem Area
grid<-grid%>%
  st_join(LME, join = st_within)
saveRDS(grid, "Data/MODIS/grid_adfg_nmfs_lme.RDS")

#Ecosystem subarea
grid<-grid%>%
  st_join(ESR, join = st_within)
saveRDS(grid, "Data/MODIS/grid_adfg_nmfs_lme_esr.RDS")

#BSIERP
grid<-grid%>%
  st_join(BSIERP, join = st_within)

#NBS crab
grid<-grid%>%
  st_join(nbs_crab, join = st_within)

#BS King crab
grid<-grid%>%
  st_join(bs_king_crab, join = st_within)

#BS Tanner crab
grid<-grid%>%
  st_join(bs_tanner_crab, join = st_within)
saveRDS(grid, "Data/MODIS/grid_all_boundaries.RDS")

#filter out all of that shape.x.1 etc fields
grid1<-grid%>%
  dplyr::select(lonc, latc, STAT_AREA, WATERS_COD, NMFS_REP_AREA, Ecosystem_Area, Ecosystem_Subarea, BSIERP_ID, BSIERP_Region_Name, NBS_CRAB, BS_KING, BS_TANNER, Shape_Length.x, Shape_Area.x, geometry)

```

**add depth**
We use NOAA bathymetry from the marmap package to attribute depth to points
```{r, eval=FALSE}
#Also load marmap
r.ak <- getNOAA.bathy(lon1=-180,lon2=-129,lat1=47.5,lat2=71, resolution=1)
#Positive longitudes
r.ak_w <- getNOAA.bathy(lon1=167,lon2=180,lat1=47.5,lat2=71, resolution=1)
#converting to a raster
r.ak2 <- marmap::as.raster( r.ak)
r.ak3 <- marmap::as.raster( r.ak_w)

r.ak4<-raster::merge(r.ak2, r.ak3)


#merge combined raster and remove positive depths
grid2<-grid1%>%mutate(depth=round(raster::extract(r.ak4,cbind(lonc,latc),method="bilinear"),0))  %>% filter(depth<=0) 
###QA

summary(grid2$depth)

#plot
ggplot()+
  geom_tile(data=grid2, aes(x=ifelse(lonc>0, lonc, lonc+360), y=latc, fill=depth, color=depth))

```

**save as csv**

```{r, eval=FALSE}
#remove spatial geometry fields
modis_grid<-grid2%>%data.frame()%>%dplyr::select(!c(Shape_Length.x, Shape_Area.x, geometry))
#export
write.csv(modis_grid, "Data/MODIS/modis_chl_spatial_lookup.csv")
```

**test**
Make sure that it works to merge this with the data

```{r}
test_join<-mod_2003%>%inner_join(modis_grid, by=c("lonc"="lonc", "latc"="latc"))
#attribution looks good
ggplot()+
  geom_point(data=test_join%>%filter(month==6), aes(x=ifelse(lonc>0, lonc, lonc+360), y=latc, color=NMFS_REP_AREA ))
#zoom in on small area

ggplot()+
  geom_point(data=test_join%>%filter(month==6), aes(x=ifelse(lonc>0, lonc, lonc+360), y=latc ))+
  xlim(c(179,181))+ylim(c(55,56))
```

## Append strata to chla data
Merge for each year
```{r}
#function to merge
chla_merge<-function(x){
  x%>%inner_join(modis_grid, by=c("lonc"="lonc", "latc"="latc"))
}

lapply(2005:2022, function(x) readRDS(paste0("Data/MODIS/ESR/mod_",x,".RDS"))%>%
                                        chla_merge()%>%
                                        saveRDS(paste0("Data/MODIS/ESR/mod_",x,"_strata.RDS")))
       
```

## Append new 2022 data

```{r}
#load modis grid
modis_grid<-read.csv("Data/MODIS/modis_chl_spatial_lookup.csv")%>%
  dplyr::select(!X)
#June
#bind E and W Jun
jun<-tidy_chl("Data/MODIS/ESR/mod_jun_2022_e.nc")%>%
  bind_rows(tidy_chl("Data/MODIS/ESR/mod_jun_2022_w.nc"))
  #append spatial strata
jun<-chla_merge(jun)
#load old 2022 data
mod2022<-readRDS("Data/MODIS/ESR/mod_2022_strata.RDS")
#combine
mod2022<-mod2022%>%bind_rows(jun)
#save
saveRDS(mod2022, "Data/MODIS/ESR/mod_2022_strata.RDS")

#compare 2022 and 2021 for merging
mod2022<-readRDS("Data/MODIS/ESR/mod_2022_strata.RDS")
mod2021<-readRDS("Data/MODIS/ESR/mod_2021_strata.RDS")
#nmfs area needs to be chr in 2022
mod2022$NMFS_REP_AREA<-as.character(mod2022$NMFS_REP_AREA)
saveRDS(mod2022, "Data/MODIS/ESR/mod_2022_strata.RDS")
```


#Pull aleutians for NOEL
```{r}
#function to filter aleutions
chla_ai<-function(x){
  x%>%filter(Ecosystem_Area=="Aleutian Islands")
}

#bind rows with other years
mod_ai<-lapply(2003:2022, function(x) readRDS(paste0("Data/MODIS/ESR/mod_",x,"_strata.RDS"))%>%
                     chla_ai())%>%
  bind_rows()

mod_ai<-mod_ai%>%
  dplyr::select(!c(lonc, latc, BSIERP_ID, BSIERP_Region_Name, NBS_CRAB, BS_KING, BS_TANNER))

saveRDS(mod_ai, "Data/MODIS/ESR/mod_ai.RDS")

#save as csv
readRDS("Data/MODIS/ESR/mod_ai.RDS")%>%
write.csv("Data/MODIS/ESR/mod_ai.csv", row.names=FALSE)
```

## compare modis with previous
```{r}
modis_old<-readRDS("Data/MODIS/merged_8day_2003_2021_Aleutians.RDS")

modis_new<-readRDS("Data/MODIS/ESR/mod_ai.RDS")

modis_old%>%
  filter(Ecosystem_sub=="Western Aleutians")%>%
  mutate(year=year(date))%>%
  group_by(year)%>%
  summarise(oldchla=round(mean(chlorophyll),3))%>%
  bind_cols(modis_new%>%
  filter(Ecosystem_Subarea=="Western Aleutians")%>%
  mutate(year=year(date))%>%
  group_by(year)%>%
  summarise(newchla=round(mean(chlorophyll),3)))


```

# Pull BS for Jens
```{r}
#function to filter bs
chla_bs<-function(x){
  x%>%filter(Ecosystem_Area=="Eastern Bering Sea")
}

#bind rows with other years
mod_bs<-lapply(2003:2022, function(x) readRDS(paste0("Data/MODIS/ESR/mod_",x,"_strata.RDS"))%>%
                     chla_bs())%>%
  bind_rows()


saveRDS(mod_bs, "Data/MODIS/ESR/mod_bs.RDS")

```


# Pull GOA for Jeanette
```{r}
#function to filter goa
chla_goa<-function(x){
  x%>%filter(Ecosystem_Area=="Gulf of Alaska")
}

#bind rows with other years
mod_goa<-lapply(2003:2022, function(x) readRDS(paste0("Data/MODIS/ESR/mod_",x,"_strata.RDS"))%>%
                     chla_goa())%>%
  bind_rows()

mod_goa<-mod_goa%>%
  dplyr::select(!c(lonc, latc, BSIERP_ID, BSIERP_Region_Name, NBS_CRAB, BS_KING, BS_TANNER))

saveRDS(mod_goa, "Data/MODIS/ESR/mod_goa.RDS")

```


## Calculate April 19 for GOA
April 16-19 daily chl are available from MODIS. Here I download them and compare with results excluding these data. 
```{r}
#east of dateline
download.file(url = "https://coastwatch.pfeg.noaa.gov/erddap/griddap/erdMH1chla1day.nc?chlorophyll%5B(2022-04-16T12:00:00Z):(2022-04-19T12:00:00Z)%5D%5B(69.02083):(46.02083)%5D%5B(-179.9792):(-129.9792)%5D&.draw=surface&.vars=longitude%7Clatitude%7CsstMasked&.colorBar=%7C%7C%7C%7C%7C&.bgColor=0xffccccff", method = "libcurl", mode="wb",destfile = "Data/MODIS/ESR/mod_apr1day_2022_e.nc")
#west
download.file(url = "https://coastwatch.pfeg.noaa.gov/erddap/griddap/erdMH1chla1day.nc?chlorophyll%5B(2022-04-16T12:00:00Z):(2022-04-19T12:00:00Z)%5D%5B(56.5):(46.5)%5D%5B(167):(179.9792)%5D&.draw=surface&.vars=longitude%7Clatitude%7CsstMasked&.colorBar=%7C%7C%7C%7C%7C&.bgColor=0xffccccff", method = "libcurl", mode="wb",destfile = "Data/MODIS/ESR/mod_apr1day_2022_w.nc")

#Merge east and west april 19
april <-tidy_chl("Data/MODIS/ESR/mod_apr1day_2022_e.nc")%>%
               bind_rows(tidy_chl("Data/MODIS/ESR/mod_apr1day_2022_w.nc"))

table(april$date)


#average
april<-april%>%group_by(latitude, longitude)%>%
  summarize(chla=mean(chlorophyll))%>%
  mutate(time=1650369600,
         date=as_datetime(time),
         lonc=as.numeric(ifelse(longitude<0, #ifelse statement keeps +- lons the same length
                                substr(longitude,1,8),
                                substr(longitude,1,7))), 
         latc=as.numeric(substr(latitude,1,6)),
         chlorophyll=round(chla,3),
         year=year(date),
         month=month(date))%>%
  dplyr::select(chlorophyll, everything())


april<-chla_merge(april)
april$NMFS_REP_AREA<-as.character(april$NMFS_REP_AREA)

saveRDS(april, "Data/MODIS/ESR/mod_apr19.RDS")

#provide to Noel
april<-readRDS("Data/MODIS/ESR/mod_apr19.RDS")
  filter(ecosystem)
```

