---
title: "chla-indicators-2022"
author: "Matt Callahan"
date: '2022-08-31'
output: html_document
---

## Notes
1) 2022 Data downloading and appending of spatial strata are in the ESR-data-2022.RMD file

2) Indicator calculation and validation of previous years is done in the chla-indicator-comparison-calc.RMD

3) This file contains brief QAQC on the already questionable MODIS data for 2022 and calculates the indicators for 2022. Previous values will be used for 2003-2021.

## QA of 2022 MODIS
```{r}
require(tidyverse)

#AMJ... 12 seems high
mod2022%>%  filter(BSIERP_Region_Name%in%c("St. Matthew",
                          "Pribilofs"
                          ) & 
           month%in%c(4:6) 
         ) %>%
  group_by(date)%>%
  summarise(MODIS_VALUE=mean(chlorophyll,na.rm=TRUE))#%>%
  ggplot()+
  geom_point(aes(x=date, y=MODIS_VALUE))

#BBRKC
 mod2022%>%  
     filter(BS_KING=="BBRKC" & month%in%c(4:6)) %>%
  group_by(date)%>%
  summarise(MODIS_VALUE=mean(chlorophyll,na.rm=TRUE))#%>%
  ggplot()+
  geom_point(aes(x=date, y=MODIS_VALUE)) 
 
 #SMBKC
  mod2022%>%  
  filter(BS_KING=="StMattBKC" & month%in%c(4:6)) %>%
  group_by(date)%>%
  summarise(MODIS_VALUE=mean(chlorophyll,na.rm=TRUE))%>%
  ggplot()+
  geom_point(aes(x=date, y=MODIS_VALUE)) 
```
# calculate missing values
we can interpolate april 11 and 19... but since we don't first average by date before calculating indicators, this isn't particularly helpful. 

```{r}
library(zoo)
#AMJ
hm<-c(0.55,NA,NA,1.50)
na.approx(hm)

#BBKC
ba<-c(0.45,NA,NA,2.33)
na.approx(ba)

```


## Pull previous years for BS
Using the same function we used for the Aleutians...

```{r}
#pull bering sea
#function to filter EBS
chla_bs<-function(x){
  x%>%filter(Ecosystem_Area=="Eastern Bering Sea")
}

#bind rows with other years
mod_ebs2<-lapply(2003:2021, function(x) readRDS(paste0("Data/MODIS/ESR/mod_",x,"_strata.RDS"))%>%
                     chla_bs())%>%
  bind_rows()
```



## AMJ_Chlorophylla_Biomass_SEBS_Satellite
This confirms that the newly downloaded MODIS data matches previously calculated indicators. 
I could have used the old mod_ebs from Jordan but since I supplied this version to Noel for the Aleutians I wanted to confirm it.

```{r}
#define indicator
indicator<-"AMJ_Chlorophylla_Biomass_SEBS_Satellite"
#MODIS

#recalculate indicator
mod_AMJ_SEBS <- mod_ebs %>%
  mutate(month=month(date),
         YEAR=year(date)) %>%
  filter(bsierp_name%in%c("St. Matthew",
                          "Pribilofs"
                          ) & 
           month%in%c(4:6) 
         ) %>%
  group_by(YEAR) %>%
  summarise(MODIS_VALUE=mean(chlorophyll,na.rm=TRUE))

#recalculate indicator
mod_AMJ_SEBS2 <- mod_ebs2 %>%
  mutate(YEAR=year)%>%
  filter(BSIERP_Region_Name%in%c("St. Matthew",
                          "Pribilofs"
                          ) & 
           month%in%c(4:6) 
         ) %>%
  group_by(YEAR) %>%
  summarise(MODIS_VALUE=mean(chlorophyll,na.rm=TRUE))#%>%

#join with previous indicator and compare
temp<-pull_ind(indicator, mod_AMJ_SEBS2)
 
```

The following chunk calculates the indicator for 2022, binds to the 2003-2021 data, and saves as a csv.
```{r}
#recalculate indicator
mod_AMJ_SEBS_2022 <- mod2022 %>%
  mutate(YEAR=year)%>%
  filter(BSIERP_Region_Name%in%c("St. Matthew",
                          "Pribilofs"
                          ) & 
           month%in%c(4:6) 
         ) %>%
  group_by(YEAR) %>%
  summarise(MODIS_VALUE=mean(chlorophyll,na.rm=TRUE))

mod_AMJ_SEBS2<-mod_AMJ_SEBS2%>%
  bind_rows(mod_AMJ_SEBS_2022)

mod_AMJ_SEBS2$INDICATOR_NAME<-indicator
mod_AMJ_SEBS2<-mod_AMJ_SEBS2%>%
  mutate(DATA_VALUE=MODIS_VALUE)%>%
  dplyr::select(c(YEAR, INDICATOR_NAME, DATA_VALUE))

write.csv(mod_AMJ_SEBS2, "2022_indicators/AMJ_CHLA_SEBS.csv", row.names=FALSE)
```



## Spring_Chlorophylla_Biomass_SEBS_Inner_Shelf_Satellite
I'm just going to calculate these 2.
```{r}
#define indicator
indicator<-"Spring_Chlorophylla_Biomass_SEBS_Inner_Shelf_Satellite"
#MODIS
#recalculate indicator
mod_BB <- mod_ebs2 %>%
  mutate(YEAR=year) %>%
  filter(BS_KING=="BBRKC" & month%in%c(4:6)) %>%
  group_by(YEAR) %>%
  summarise(DATA_VALUE=mean(chlorophyll,na.rm=TRUE))%>%
  mutate(INDICATOR_NAME=indicator)%>%
  dplyr::select(c(YEAR, INDICATOR_NAME, DATA_VALUE))

#recalculate for 2022
bb22 <- mod2022 %>%
  mutate(YEAR=year)%>%
  filter(BS_KING=="BBRKC" & month%in%c(4:6)) %>%
  group_by(YEAR) %>%
  summarise(DATA_VALUE=mean(chlorophyll,na.rm=TRUE))%>%
  mutate(INDICATOR_NAME=indicator)%>%
  dplyr::select(c(YEAR, INDICATOR_NAME, DATA_VALUE))

mod_BB<-mod_BB%>%
  bind_rows(bb22)

write.csv(mod_BB, "2022_indicators/SPRING_CHLA_SEBS_IS.csv", row.names=FALSE)
```

## Spring_Chlorophylla_Biomass_SMBKC_Satellite

```{r}
#define indicator
indicator<-"Spring_Chlorophylla_Biomass_SMBKC_Satellite"
#MODIS
#recalculate indicator
mod_smbkc <- mod_ebs2 %>%
  mutate(YEAR=year) %>%
  filter(BS_KING=="StMattBKC" & month%in%c(4:6)) %>%
  group_by(YEAR) %>%
  summarise(DATA_VALUE=mean(chlorophyll,na.rm=TRUE))%>%
  mutate(INDICATOR_NAME=indicator)%>%
  dplyr::select(c(YEAR, INDICATOR_NAME, DATA_VALUE))

#recalculate indicator for 2022
sm22 <- mod2022 %>%
  mutate(YEAR=year) %>%
  filter(BS_KING=="StMattBKC" & month%in%c(4:6)) %>%
  group_by(YEAR) %>%
  summarise(DATA_VALUE=mean(chlorophyll,na.rm=TRUE))%>%
  mutate(INDICATOR_NAME=indicator)%>%
  dplyr::select(c(YEAR, INDICATOR_NAME, DATA_VALUE))

mod_smbkc<-mod_smbkc%>%
  bind_rows(sm22)

write.csv(mod_smbkc, "2022_indicators/SPRING_CHLA_SMBKC.csv", row.names=FALSE)
```

## Groundfish indicators
Here I calculate 2022 groundfish indicators for the GOA. Jens Nielson will calculate the BS

Spring_Chlorophylla_Peak_EGOA_Satellite
Spring_Chlorophylla_Peak_WCGOA_Satellite
Spring_Chlorophylla_Biomass_EGOA_Satellite
Spring_Chlorophylla_Biomass_WCGOA_Satellite

 
##GOA    

```{r}
#pull bering sea
#function to filter EBS
chla_goa<-function(x){
  x%>%filter(NMFS_REP_AREA %in% c(610, 620, 630, 640, 650))
}

#bind rows with other years
mod_goa<-lapply(2003:2022, function(x) readRDS(paste0("Data/MODIS/ESR/mod_",x,"_strata.RDS"))%>%
                     chla_goa())%>%
  bind_rows()
```



## Spring_Chlorophylla_Biomass_GOA_Satellite

This was originally Spring_Chlorophylla_Biomass_EGOA_Satellite. At the request of the sablefish ESP team we expanded the extent to gulf-wide and removed the depth filter.

```{r}
#define indicator
indicator<-"Spring_Chlorophylla_Biomass_GOA_Satellite"

#calculate indicator value
mod_SPR_GOA_B <- mod_goa %>%
  mutate(YEAR=year) %>%
  filter(NMFS_REP_AREA %in% c(610, 620, 630, 640, 650) & month %in% c(4:6) & depth<(-10)) %>%
  group_by(YEAR) %>%
  summarise(DATA_VALUE=mean(chlorophyll,na.rm=TRUE))%>%
  mutate(INDICATOR_NAME=indicator)%>%
  dplyr::select(c(YEAR, INDICATOR_NAME, DATA_VALUE))

write.csv(mod_SPR_GOA_B, "2022_indicators/Spring_Chlorophylla_Biomass_GOA_Satellite.csv", row.names=FALSE)
```

## Spring_Chlorophylla_Peak_GOA_Satellite

This was originally Spring_Chlorophylla_Biomass_EGOA_Satellite. At the request of the sablefish ESP team we expanded the extent to gulf-wide and removed the depth filter.

```{r}
#define indicator
indicator<-"Spring_Chlorophylla_Peak_GOA_Satellite"

#calculate indicator value
mod_SPR_GOA_P <- mod_goa %>%
  filter(NMFS_REP_AREA %in% c(610, 620, 630, 640, 650) & month %in% c(4:6) & depth<=(-10)) %>%
  group_by(date) %>%
  summarise(WEEKLY_VALUE=mean(chlorophyll,na.rm=TRUE))%>%
  mutate(YEAR=year(date))%>%
  group_by(YEAR)%>%
  slice(which.max(WEEKLY_VALUE))%>%
  mutate(DATA_VALUE=yday(date))%>%
  mutate(INDICATOR_NAME=indicator)%>%
  dplyr::select(c(YEAR, INDICATOR_NAME, DATA_VALUE))

write.csv(mod_SPR_GOA_P, "2022_indicators/Spring_Chlorophylla_Peak_GOA_Satellite.csv", row.names=FALSE)
```


## Spring_Chlorophylla_Biomass_WCGOA_Satellite

I used the same filters Jordan included in his github code, the values are close, but do not match perfectly.

```{r, warning=FALSE, message=FALSE}
#define indicator
indicator<-"Spring_Chlorophylla_Biomass_WCGOA_Satellite"

#calculate indicator value
mod_SPR_WCGOA_B <- mod_goa %>%
  mutate(YEAR=year) %>%
  filter(NMFS_REP_AREA %in% c(610, 620, 630) & month==5 & depth<(-10) & depth>(-200)) %>%
  group_by(YEAR) %>%
  summarise(DATA_VALUE=mean(chlorophyll,na.rm=TRUE))%>%
  mutate(INDICATOR_NAME=indicator)%>%
  dplyr::select(c(YEAR, INDICATOR_NAME, DATA_VALUE))

#write csv
write.csv(mod_SPR_WCGOA_B, "2022_indicators/Spring_Chlorophylla_Biomass_WCGOA_Satellite.csv", row.names=FALSE)

```

## Spring_Chlorophylla_Peak_WCGOA_Satellite

```{r}
#define indicator
indicator<-"Spring_Chlorophylla_Peak_WCGOA_Satellite"
#MODIS
#recalculate indicator
mod_SPR_WCGOA_P <- mod_goa %>%
  filter(NMFS_REP_AREA %in% c(610, 620, 630) & month %in% c(4:6) & depth<=(-10) & depth>=(-200)) %>%
  group_by(date) %>%
  summarise(WEEKLY_VALUE=mean(chlorophyll,na.rm=TRUE))%>%
  mutate(YEAR=year(date))%>%
  group_by(YEAR)%>%
  slice(which.max(WEEKLY_VALUE))%>%
  mutate(DATA_VALUE=yday(date))%>%
  mutate(INDICATOR_NAME=indicator)%>%
  dplyr::select(c(YEAR, INDICATOR_NAME, DATA_VALUE))


#write csv
write.csv(mod_SPR_WCGOA_P, "2022_indicators/Spring_Chlorophylla_Peak_WCGOA_Satellite.csv", row.names=FALSE)
```

## quickly plot chla indicators
```{r}
ggplot()+
  geom_line(data=mod_SPR_WCGOA_P, aes(x=YEAR, y=DATA_VALUE))+
  geom_line(data=mod_SPR_GOA_P, aes(x=YEAR, y=DATA_VALUE), color="red")+
  ggtitle("Peak chla: red=GOA, black=WCGOA")

ggplot()+
  geom_line(data=mod_SPR_WCGOA_B, aes(x=YEAR, y=DATA_VALUE))+
  geom_line(data=mod_SPR_GOA_B, aes(x=YEAR, y=DATA_VALUE), color="red")+
  ggtitle("Chla biomass: red=GOA, black=WCGOA")
```

